---
title: "Airline Passenger Satisfaction"
author: "Sergio E. Sagastume"
date: "2025-06-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This project explores airline passenger satisfaction survey data to understand what factors influence how satisfied travelers feel. By using data from Kaggle, I aim to identify patterns, build a model, and provide useful recommendations to improve airline services.

# Scenario

Airlines rely heavily on feedback to improve customer experience and retain loyal passengers. With this dataset, I can look at survey ratings related to booking, boarding, comfort, and service, and explore how they connect to overall satisfaction.

# Ask

The key questions for this analysis are:

1. What factors are most related to passenger satisfaction?
2. How does satisfaction vary by customer type, travel class, or travel purpose?
3. Can I predict satisfaction using this data?
4. What can airlines do to increase passenger satisfaction?

# Prepare

**Data Source**: Kaggle – Airline Passenger Satisfaction  
Dataset: [Airline Passenger Satisfaction](https://www.kaggle.com/datasets/teejmahal20/airline-passenger-satisfaction)
```{r prepare, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(caret)

# Load and clean
train <- read_csv("Airline Satisfaction Prediction Analysis/train.csv") %>%
  clean_names() %>%
  select(-id) %>%
  mutate(across(c(gender, customer_type, type_of_travel, class, satisfaction), as.factor),
         arrival_delay_in_minutes = replace_na(arrival_delay_in_minutes, 0),
         satisfaction_binary = ifelse(satisfaction == "satisfied", 1, 0))
```

# Process
```{r}
# Check for missing values
colSums(is.na(train))

# Final structure
glimpse(train)
```

# Analyze
```{r}
# Split and model
set.seed(123)
index <- createDataPartition(train$satisfaction_binary, p = 0.8, list = FALSE)
train_set <- train[index, ]
test_set <- train[-index, ]

model <- glm(satisfaction_binary ~ gender + customer_type + type_of_travel + class +
               flight_distance + inflight_wifi_service + ease_of_online_booking +
               online_boarding + seat_comfort + inflight_entertainment +
               cleanliness + departure_delay_in_minutes + arrival_delay_in_minutes,
             data = train_set, family = binomial)

# Predict and evaluate
pred <- ifelse(predict(model, test_set, type = "response") > 0.5, 1, 0)
confusionMatrix(factor(pred), factor(test_set$satisfaction_binary))
```

# Share
```{r}
# Satisfaction breakdown
ggplot(train, aes(satisfaction)) +
  geom_bar(fill = "#2c3e50") +
  labs(title = "Passenger Satisfaction",
       caption = "Source: Airline Passenger Satisfaction dataset on Kaggle") +
  theme_minimal()

# By customer type
ggplot(train, aes(customer_type, fill = satisfaction)) +
  geom_bar(position = "fill") +
  labs(title = "Satisfaction by Customer Type") +
  theme_minimal()

# Service ratings
train %>%
  select(satisfaction, online_boarding, seat_comfort, inflight_wifi_service,
         inflight_entertainment, cleanliness) %>%
  pivot_longer(-satisfaction, names_to = "service", values_to = "rating") %>%
  group_by(satisfaction, service) %>%
  summarise(avg_rating = mean(rating), .groups = "drop") %>%
  ggplot(aes(service, avg_rating, fill = satisfaction)) +
  geom_col(position = "dodge") +
  labs(title = "Average Ratings by Satisfaction",
       x = "Service", y = "Avg Rating") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Act
```{r}
# Top predictors
coef_summary <- summary(model)$coefficients
sorted <- sort(coef_summary[, "Estimate"], decreasing = TRUE)

head(sorted, 5)
tail(sorted, 5)

# Recommendations
cat("Key Recommendations:\n")
cat("- Improve inflight services, especially entertainment and seat comfort.\n")
cat("- Focus loyalty efforts on business travelers and frequent flyers.\n")
cat("- Expand Business class benefits, as it links strongly with satisfaction.\n")
cat("- Enhance digital tools for easier boarding and booking.\n")
cat("- Work to reduce delays, as they negatively impact satisfaction.\n\n")

cat("Source: Airline Passenger Satisfaction dataset on Kaggle\n")
```

q()
